---
title: "Partial correlation p value"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview
**Partial correlation** is the correlation between X and Y after partialling out the effect of Z(s). Partialling out is sometimes described as "controlling for", "regressing out", "adjusting for", "accounting for", etc. 

This value can be computed as the correlation between the residuals of X after regressing onto Z, and the residuals of Y after regressing onto Z. 


# Issue
Simply running `cor(x.resid, y.resid)`, p-value is slightly different than the p-value from partial correlation functions such as `ppcor::pcor.test`j, albeit only slightly.

## Here is an example:

Load data `mtcars`
```{r data}
df <- mtcars
```


Calculate partial correlation using `pcor.test` between `mpg` and `cyl`, partialling out `hp`

```{r}
prr <- ppcor::pcor.test(x = df$mpg, y = df$cyl, z = df$hp)
```

partial r = `r prr$estimate`, p = `r prr$p.value`

Lets compare the above to using `cor()` with the residuals of `mpg` and `cyl` after partialling out `hp`

```{r}
# Getting `residuals` from `lm()`
x.resid <- resid(lm(mpg ~ hp, data=df))
y.resid <- resid(lm(cyl ~ hp, data=df))

rr <- cor.test(x.resid, y.resid)
```

partial r = `r rr$estimate`, p = `r rr$p.value`

**While the r values from both methods are the same, but the p-value is slightly different. WHY?**

# Reasons

The degree of freedom(df) in a simple correlation is `n-2`.

However, the df in partial correlation is `n-2-g`, where `g` is the number of variables being partialled out. 

In the above example, only `hp` was partialled out, therefore `g=1`. Because `cor()` assumes the df of 2, the p-value is incorrect (albeit only slightly in this case; however, this can become more problematic as more variables are partialled out [i.e., greater `g`]). 

# Sanity Check - Manually calculate p-value with correct df

```{r}
par.r <- as.numeric(rr$estimate)
n <- nrow(df)
g <- 1
degf <- n-2-g

tstats <- par.r*sqrt((degf)/(1-par.r^2))
pv <- 2*pt(abs(tstats), degf, lower.tail = F)
```

Manually calculate the p-value using the correct df (`r degf`) yield a p-value of `r pv`, the same as the output from `pcor.test`.


